Type theories as quotient inductive-recursive types

### 2025-02-18 tex/main.tex
Finish methods

One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\section{Complete definitions}\label{appendix:full-def}
We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.

\paragraph*{Substitution calculus with a base type $\UU$}
\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set \\
  [\blank]_{\Ty}\blank & : \implicit{\Gamma, \Delta}\; && \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
  \emptyctx & : && \Ctx \\
  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx \\
  \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;([ \sigma ]_{\Ty}\,A) \to \Sub\;\Gamma,(\Delta, A) \\
  \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\;\sigma ]_{\Ty}\, A) \\
  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A) \\
  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
\end{alignat*}
\begin{alignat*}{5}
  & [ \idS ]_{\Ty}\,               && A           && \reduce A \\
  & [ \sigma ; \tau ]_{\Ty}\,      && A           && \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
  & [ \pi_1(\sigma, t) ]_{\Ty}\,   && A           && \reduce [\sigma]_\Ty\,A \\
  & [ \pi_1(\sigma; \tau) ]_{\Ty}\,&& A           && \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
  & [ \sigma ]_{\Ty}\,             && \UU         && \reduce \UU
\end{alignat*}
\begin{alignat*}{5}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  [\idS]t         & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
  [;]t            & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
\end{alignat*}

\paragraph*{Universe of small types}
\begin{alignat*}{5}
  [\blank]_{\Tm}\blank            & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\,A) \\ 
 [ \idS ]_{\Tm}\,t                & \reduce t \\
 [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
 [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
 [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
 [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}

@@

\begin{alignat*}{5}
  \nameless & : [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
\end{alignat*}

\paragraph*{$\Pi$-types}
\begin{alignat*}{5}
  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
\end{alignat*}
\begin{alignat*}{5}
  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
\idS                \uparrow A  & \reduce \idS \\
\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
\end{alignat*}
\begin{alignat*}{5}
  \Pi             & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
\begin{alignat*}{5}
  \nameless       &: && [ \sigma ]_{\Ty}\, (\Pi\,A\,B) && {} \reduce {} && \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B}) \\
  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& [ \sigma ] \,(\lambda\,t) && {} =^{\Tm\;\Gamma\;([\sigma]_{\Ty}(\Pi\,A\,B)} {} && \lambda ([\sigma \uparrow A ]\,t)\\
  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
\end{alignat*}

\paragraph*{Extensional identity types}

\begin{alignat*}{5}
  \mathsf{Id} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
  \nameless & : && [ \sigma ]_{\Ty}\, \idtype[a]{t}{u} \reduce \idtype[{[\sigma]_{\Tm}\,a}]{[\sigma]_{\Tm}\,t}{[\sigma]_{\Tm}\,u} \\
  \mathsf{refl} & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\mathsf{Id}\;a\;t\;t) \\
  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
\end{alignat*}

\paragraph*{Elimination principle}

@@

\begin{alignat*}{5}
  \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
  ;\mathord{\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
  \pi_1,\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
  \pi_1;\mathord{\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
  \emptysub^M_{\Sub} \mathord{\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
  ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
  \pi_1\idS\mathord{\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
  \pi_1\pi_1\mathord{\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
\end{alignat*}

>>>>

One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define$[\blank]_{\Ty} \blank : \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma$ recursively as follows:
\begin{alignat}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\section{Complete definitions}\label{appendix:full-def}

\paragraph*{Quotient inductive-inductive-recursive types}
We give the QIIR definition for type theory with a universe of small types, $\Pi$-types, and extensional identity types in one place below.

\subparagraph*{Substitution calculus with a base type $\UU$}
\begin{alignat*}{3}
  \Ctx      & : && \Set                   \\
  \Ty       & : && \Ctx \to \Set          \\
  \Sub      & : && \Ctx \to \Ctx \to \Set \\
  \Tm       & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Set
\end{alignat*}
\begin{alignat*}{5}
  \emptyctx & : && \Ctx \\
  \blank,\blank & : && (\Gamma : \Ctx) \to \Ty\;\Gamma \to \Ctx
\end{alignat*}
\begin{alignat*}{5}
  [\blank]_{\Ty}\blank & : \implicit{\Gamma, \Delta}\; && \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
  \UU     & : \implicit{\Gamma}\; && \Ty\; \Gamma
\end{alignat*}
\begin{alignat*}{5}
  \emptysub & : \implicit{\Gamma} && \Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;([ \sigma ]_{\Ty}\,A) \to \Sub\;\Gamma,(\Delta, A) \\
  \idS & : \implicit{\Gamma} && \Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta} && \Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta \\
  \pi_1 & : \implicit{\Gamma, \Delta, A}\; && \Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\; && (\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\;\sigma ]_{\Ty}\, A) \\
  [\blank] \blank & : \implicit{\Gamma,\Delta, A} && (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\, A)
\end{alignat*}
\begin{alignat*}{5}
  & [ \idS ]_{\Ty}\,               && A           && \reduce A \\
  & [ \sigma ; \tau ]_{\Ty}\,      && A           && \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \\
  & [ \pi_1(\sigma, t) ]_{\Ty}\,   && A           && \reduce [\sigma]_\Ty\,A \\
  & [ \pi_1(\sigma; \tau) ]_{\Ty}\,&& A           && \reduce [\sigma]_\Ty ([\pi_1\tau]_\Ty \,A) \\
  & [ \sigma ]_{\Ty}\,             && \UU         && \reduce \UU
\end{alignat*}
\begin{alignat*}{5}
  ;\text{-}\mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\sigma ; \idS_{\Delta}} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma}\; && {\idS_{\Gamma} ; \sigma} && =^{\Sub\;\Gamma\;\Delta} && {\sigma} \\
  ;\text{-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\;\Gamma\;\Theta} &&  \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t}\; &&\sigma ; (\tau , t) && =^{\Sub\;\Gamma\;(\Theta, A)} &&  (\sigma ; \tau) , [ \sigma ] t \\
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\;\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\;\Gamma\; A} &&  t \\
  \emptyctx\eta   & : \implicit{\Gamma, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;\emptyctx} & \emptysub \\
  \pi\eta         & : \implicit{\Gamma, A, \sigma}\; && \sigma                   && =^{\Sub\;\Gamma\;(\Delta, A)} &&  (\pi_1 \sigma, \pi_2 \sigma) \\
  [\idS]          & : \implicit{\Gamma, A, t}\; && {[\idS]\;t}          && =^{\Tm\;\Gamma\;A} && t \\
  [;]             & :\implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; && {[\sigma ; \tau]\;t} && =^{\Tm\;\Gamma\;[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
\end{alignat*}

\subparagraph*{Universe of small types}
\begin{alignat*}{5}
  [\blank]_{\Tm}\blank            & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]_{\Ty}\,A) \\ 
 [ \idS ]_{\Tm}\,t                & \reduce t \\
 [ \sigma ; \tau ]_{\Tm}\,t       & \reduce [ \sigma ]_{\Tm}\;([ \tau ]_{\Tm}\;t) \\
 [ \pi_1(\sigma, t) ]_{\Tm}\,t    & \reduce [\sigma]_\Tm\,t \\
 [ \pi_1(\sigma; \tau) ]_{\Tm}\,t & \reduce [\sigma]_\Tm\; ([\pi_1\tau]_\Tm \,t) \\
 [ \sigma ]_{\Tm}\,t              & \reduce [ \sigma ]\,t, \quad \text{otherwise}
\end{alignat*}

@@

\begin{alignat*}{5}
  [ \sigma ]_{\Ty}\,  && (\El\, u) \reduce \El\,([\sigma]_{\Tm}{u})
\end{alignat*}

\subparagraph*{$\Pi$-types}
\begin{alignat*}{5}
  \blank^+ & : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\;\Gamma\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
  \sigma^+ & \defeq (\pi_1\idS; \sigma), \pi_2\sigma
\end{alignat*}
\begin{alignat*}{5}
  \blank\uparrow\blank & : \implicit{\Gamma, \Delta}\;(\sigma : \Sub\;\Gamma\;\Delta) (A : \Ty\;\Delta) \to \Sub\;(\Gamma, [\sigma]_{\Ty}\;A)\;(\Delta, A) \\
\idS                \uparrow A  & \reduce \idS \\
\sigma ; \tau       \uparrow A  & \reduce (\sigma \uparrow \sub{\tau}{A} ) ; (\tau \uparrow A) \\
\pi_1(\sigma, t)    \uparrow A  & \reduce \sigma \uparrow A \\
\pi_1(\sigma; \tau) \uparrow A  & \reduce \sigma \uparrow (\sub{\pi_1 \tau}{A}) ; (\pi_1 \tau \uparrow A) \\
\sigma              \uparrow A  & \reduce \sigma^+,\quad  \text{otherwiese}
\end{alignat*}
\begin{alignat*}{5}
  \Pi             & : \implicit{\Gamma}\; && (A : \Ty\;\Gamma) \to \Ty\;(\Gamma, A) \to \Ty\;\Gamma \\
  \lambda         &: \implicit{\Gamma, A, B} \; && \Tm\;(\Gamma, A)\;B \to \Tm\;\Gamma(\Pi\,A\,B) \\
  \mathsf{app}    &: \implicit{\Gamma, A, B} \; && \Tm\;\Gamma\;(\Pi\;A\;B) \to \Tm\;(\Gamma, A)\;B
\end{alignat*}
\begin{alignat*}{5}
  [ \sigma ]_{\Ty}\, (\Pi\,A\,B) \reduce \Pi\,(\sub{\sigma}{A})\,(\sub{\sigma\uparrow A}{B})
\end{alignat*}
\begin{alignat*}{5}
  []\lambda       & : \implicit{\Gamma, \Delta, \sigma, A, t}\;&& [ \sigma ] \,(\lambda\,t) && {} =^{\Tm\;\Gamma\;([\sigma]_{\Ty}(\Pi\,A\,B)} {} && \lambda ([\sigma \uparrow A ]\,t)\\
  \Pi\beta        & : \implicit{\Gamma, A, B, t}\; && \mathsf{app}\,(\lambda\,t)            && =^{\Tm\;\Gamma\;A} && t \\
  \Pi\eta         & : \implicit{\Gamma, A, B, t}\; && t                                     && =^{\Tm\;\Gamma\;A} && \lambda(\mathsf{app}\,t)
\end{alignat*}

\subparagraph*{Extensional identity types}

\begin{alignat*}{5}
  \mathsf{Id} & : \implicit{\Gamma} && (a : \Tm\;\Gamma\;\UU) (t : \Tm\;\Gamma\;(\El\,a)(u : \Tm\;\Gamma\;(\El\,a)) \to \Ty\;\Gamma \\
  \nameless & : && [ \sigma ]_{\Ty}\;(\mathsf{Id}\;a\;t\;u) \reduce \mathsf{Id}\;([\sigma]_{\Tm}\,a)\;([\sigma]_{\Tm}\,t)\;([\sigma]_{\Tm}\,u) \\
  \mathsf{refl}   & : \implicit{\Gamma, a, t}\;&&\Tm\;\Gamma\;(\mathsf{Id}\;a\;t\;t) \\
  []\mathsf{refl} & : \implicit{\Gamma, a, t}\;&& [\sigma]\,(\mathsf{refl}\;t) =^{\Tm\;\Gamma\;([\sigma](\mathsf{Id}\,a\,t\,t))} \mathsf{refl}([ \sigma ]_{\Tm}\,t) \\
  \mathsf{reflect} & : \implicit{\Gamma, a, t, u}\;&& \Tm\;\Gamma\;(\idtype[a]{t}{u}) \to t =^{\Tm\;\Gamma\;(\El\,a)} u
\end{alignat*}

\paragraph*{Elimination principles for type theories as QIIRTs}

@@

\begin{alignat*}{5}
  \idS\mathord{\uparrow}^M      & : \idS^M              && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\idS}_{\alert{[\idS]^M_{\Ty}}} {} && \idS^M \\
  \mathord{;\uparrow}^M         & : (\sigma^M ; \tau^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M [ \tau^M ]^M_{\Ty} \,A^M) ;^M (\tau^M \uparrow^M A^M)  \\
  \mathord{\pi_1,\uparrow}^M    & : \pi_1^M(\sigma^M ,^M_{\Sub} t^M) && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1,]^M_{\Ty}}}            {} && \sigma^M \uparrow^M  A^M \\
  \mathord{\pi_1;\uparrow}^M    & : \pi_1^M(\sigma^M ;^M \tau^M)     && \uparrow^M A^M  =^{\lam{B^M} \Sub^M\,(\Gamma^M ,_{\Ctx}^M B^M)\,(\Gamma^M ,_{\Ctx}^M A^M)\;\nameless}_{\alert{[\pi_1;]^M_{\Ty}}} {} && (\sigma^M \uparrow^M  [\pi_1^M\tau^M]^M_{\Ty}\,A^M);^M (\pi_1^M\tau^M \uparrow^M A^M) \\
  \mathord{\emptysub^M_{\Sub}\uparrow}^M & : \emptysub^M_{\Sub} && \uparrow^M A^M = (\emptysub^M_{\Sub})^{+M} \\
  ,\mathord{\uparrow}^M & : (\sigma^M ,_{\Sub}^M t^M)  && \uparrow^M A^M = (\sigma^M ,_{\Sub}^M t^M)^{+M} \\
  \mathord{\pi_1\idS\uparrow}^M & : (\pi_1^M\,\idS^M) && \uparrow^M A^M = (\pi_1^M\,\idS^M)^{+M} \\
  \mathord{\pi_1\pi_1\uparrow}^M & : (\pi_1^M\,(\pi_1^M\,\sigma^M)) && \uparrow^M A^M = (\pi_1^M\,(\pi_1^M\,\sigma^M))^{+M}
\end{alignat*}
